Making Sequential Programs Deterministic

### 2019-03-22 design.tex
more text

\section{Design}
\subsection{Type checking background}
The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
The hierarchy of types in this lattice defines subtyping relationships among them.
In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
\begin{figure}
    \begin{center}
        \includegraphics[scale=0.15]{lattice}
    \end{center}
    \caption{Example type hierarchy}
    \label{fig-example-lattice}
\end{figure}
In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
\begin{figure}
    \begin{verbatim}
    @A int x; @B int y; @C int z;
    x = y;    // Because types @A and @B are incomparable.
    z = x;    // Because @C is a subtype of @A.
    \end{verbatim}
\caption{Example: Invalid assingment}
\label{code-invalid1}
\end{figure}

@@

\subsection{Determinism type hierarchy}
The Determinism type system uses the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}) (We have omitted the
\<@Top> and \<@Bottom> types for brevity):
\begin{itemize}
    \item \<@NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<@OrderNonDet> indicates that
    a collection (i.e any class that is a subtype of java.util.Collection, java.util.Iterator, or java.util.Map) or an array will have the same elements in every execution, but in a
    possibly different order.  \<@OrderNonDet> may only be written on
    collections and arrays.
    \item \<@Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

@@

\subsection{Determinism}

@@

\subsection{Additional typing rules for collection elements}
The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<@NonDet>, then the type parameter may not be
\<@Det> or \<@OrderNonDet>. Although such types could exist, they are
disallowed to prevent the following bug:

@@

\subsection{Polymorphism extensions}

@@

\subsection{Differentiating binding and use}

@@

\subsection{Dataflow analysis and type refinement}
The checker framework performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
The checker framework performs dataflow analysis within method bodies and refines the types of local variables.
In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{@Det} variable \codeid{z} is sound. In the follwing example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

@@

\subsection{CLIMB-to-top and collection, array locals}

Every type system in the checker framework must define a default type annotation. This annotation will be applied
to every location that isn't explicitly annotated.

\begin{itemize}
	\item CF defaults types at CLIMB locations to top and type refines them. (steal some text from manual).
	\item CLIMB locations do not include type parameters. So this is problematic for determinism checker
	\item Example: Unannotated collection; its default would be @NonDet<@Det> which is invalid.
	\item So we default locals collections to @NonDet<@NonDet>. 
	\item This is sound but could produce a lot of FPs. Example assignment: @NonDet<@NonDet> collection = collection1.
	Type of collection1 is @Det<@Det> and it is a method argument.
	\item To avoid this, we require users to explicitly annotate local collections.
	\item Similarly for local arrays.
\end{itemize}

\subsection{Discussion on defaulting rules}

\begin{itemize}
    \item Determinism checker has \<@Det> as the usual default. Reason being we expect the program to be 
    deterministic unless specified by programmer or nondeterminism from libraries that we have annotated (Random, Collections, etc).
    \item All method parameters and return types (including constructors) are \<@PolyDet>.
    \item Reason: We want to be able to report nondeterministic behavior only where it is invalid.
    \item Example of a method with params and return types annotated as \<@PolyDet>.
    This method called from 2 places - one harmless(@Det instantiations), and another that call with \<@NonDet> args and prints the retun type.
    \item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
    We explicitly annotate the return type on constructors as
        \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
        HashMap>.
\end{itemize}

\subsection{Additional hardcoded rules in the determinism checker  for precision}

\begin{itemize}
    \item \<Set.equals()> is annotated as \<@Poly("up")>. This is sound but imprecise.
    \item Example1: \codeid{@OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>)} is deterministic
    \item Example2: \codeid{@OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>)} is not deterministic.
    \item Explain the rules under which we make Set equality deterministic.
    \item We annotate System.getProperty as \<@NonDet> because the results could vary across machines.
    \item Exceptions: line.Sepaator, file.Separator. and path.Separator.
\end{itemize}

>>>>

\section{Design}\label{design}
\subsection{Type checking background}\label{type-checking}
The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
The hierarchy of types in this lattice defines subtyping relationships among them.
In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
\begin{figure}
    \begin{center}
        \includegraphics[scale=0.15]{lattice}
    \end{center}
    \caption{Example type hierarchy}
    \label{fig-example-lattice}
\end{figure}
In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
\begin{figure}
    \begin{verbatim}
    @A int x; @B int y; @C int z;
    x = y;    // Because types @A and @B are incomparable.
    z = x;    // Because @C is a subtype of @A.
    \end{verbatim}
\caption{Example: Invalid assingment}
\label{code-invalid1}
\end{figure}

@@

\subsection{Determinism type hierarchy}\label{type-hierarchy}
The Determinism type system uses the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}) (We have omitted the
\<@Top> and \<@Bottom> types for brevity):
\begin{itemize}
    \item \<@NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<@OrderNonDet> indicates that
    a collection (i.e any class that is a subtype of java.util.Collection, java.util.Iterator, or java.util.Map) or an array will have the same elements in every execution, but in a
    possibly different order.  \<@OrderNonDet> may only be written on
    collections and arrays.
    \item \<@Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

@@

\subsection{Determinism}\label{determinism}

@@

\subsection{Additional typing rules for collection elements}\label{collection-rules}
The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<@NonDet>, then the type parameter may not be
\<@Det> or \<@OrderNonDet>. Although such types could exist, they are
disallowed to prevent the following bug:

@@

\subsection{Polymorphism extensions}\label{polymorphism}

@@

\subsection{Differentiating binding and use}\label{bindings-uses}

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
The checker framework performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
The checker framework performs dataflow analysis within method bodies and refines the types of local variables.
In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{@Det} variable \codeid{z} is sound. In the follwing example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

@@

\subsection{CLIMB-to-top and collection, array locals}\label{climb-rules}

Every type system in the checker framework must define a default type annotation. This annotation will be applied
to every location that isn't explicitly annotated. In our determinism checker, we chose \codeid{@Det} as the default annotation.
The reason for this design choice is that we expect the program to be 
deterministic unless specified by programmer or the program calls nondeterministic library methods that we have annotated 
(Random, Collections, etc).
The framework applies the CLIMB-to-top rule. This rule states that the \textit{top} qualifier 
in the hierarchy is the default for the CLIMB locations: Casts, Locals, Instanceof, and (some) iMplicit Bounds. The rationale
for this rule is as follows:

\begin{itemize}
	\item Local variables are defaulted to top because type refinement is applied to local variables. If a local variable starts as the 
	\textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
	result, a programmer rarely writes an explicit annotation on any of those locations.
	Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
	in \textit{for} statements, and \textit{catch} arguments (known as exception parameters in the Java Language Specification). 
	\textit{Exception parameters} need to have the \textit{top} type because exceptions of arbitrary qualified types can be thrown 
	and the Checker Framework does not provide runtime checks.
	\item \textit{Cast} and \textit{instanceof} types are given the same type as their argument expression. This has the same effect as 
	if they were given the \textit{top} type and then flow-sensitively refined to the type of their argument.
	\item Implicit upper bounds are defaulted to top to allow them to be instantiated in any way. If a user declared class \codeid{C<T> 
		{ ... }}, then the Checker Framework assumes that the user intended to allow any instantiation of the class, and the declaration
	is interpreted as class \codeid{C<T extends @NonDet Object> { ... }} rather than as class \codeid{C<T extends @Det Object> { ... }}. The latter would forbid instantiations such as \codeid{C<@Det String>}, or would require rewriting of code. On the other hand, if a user writes 
	an explicit bound such as class \codeid{C<T extends D> { ... }}, then the user intends some restriction on instantiation and can write a 
	qualifier on the upper bound as desired.
	This rule means that the upper bound of class \codeid{C<T>} is defaulted differently than the upper bound of class \codeid{C<T extends Object>}. 
	This may seem confusing, but it is the least bad option. The more confusing alternative would be for \codeid{Object} to be defaulted 
	differently in class \codeid{C<T extends Object>} and in an instantiation \codeid{C<Object>}, and for the upper bounds to be defaulted differently 
	in class \codeid{C<T extends Object>} and class \codeid{C<T extends Date>}.
	\item \textit{Implicit lower bounds} are defaulted to the \textit{bottom} type, again to allow maximal instantiation. Note that Java does not allow a programmer to express both the upper and lower bounds of a type, but the Checker Framework allows the programmer to specify either or both.
\end{itemize}
With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<@NonDet>.

\begin{itemize}
	\item So we default locals collections to @NonDet<@NonDet>. 
	\item This is sound but could produce a lot of FPs. Example assignment: @NonDet<@NonDet> collection = collection1.
	Type of collection1 is @Det<@Det> and it is a method argument.
	\item To avoid this, we require users to explicitly annotate local collections.
	\item Similarly for local arrays.
\end{itemize}

\subsection{Discussion on defaulting rules}\label{defaulting}

\begin{itemize}
    \item All method parameters and return types (including constructors) are \<@PolyDet>.
    \item Reason: We want to be able to report nondeterministic behavior only where it is invalid.
    \item Example of a method with params and return types annotated as \<@PolyDet>.
    This method called from 2 places - one harmless(@Det instantiations), and another that call with \<@NonDet> args and prints the retun type.
    \item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
    We explicitly annotate the return type on constructors as
        \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
        HashMap>.
\end{itemize}

\subsection{Determinism checker specific rules for precision}

\begin{itemize}
    \item \<Set.equals()> is annotated as \<@Poly("up")>. This is sound but imprecise.
    \item Example1: \codeid{@OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>)} is deterministic
    \item Example2: \codeid{@OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>)} is not deterministic.
    \item Explain the rules under which we make Set equality deterministic.
    \item We annotate System.getProperty as \<@NonDet> because the results could vary across machines.
    \item Exceptions: line.Sepaator, file.Separator. and path.Separator.
    \item discuss type on Exceptions and unsoundness due to this.
\end{itemize}
